Type-Safe Client-Server Communication in Web Applications

### 2014-02-26 paper.tex
Add some macros and a bit of text about Haste.Foreign.

\usepackage{amsmath}

@@

\authorinfo{Koen Claessen ans Anton Ekblad}
           {Chalmers University of Technology}
           {\{koen,antonek\}@chalmers.se}

@@

The runtime data representation of Haste programs is kept as close to regular
Javascript programs as possible. The numeric types are represented using the
Javascript \texttt{Number} type, which is defined as the IEEE754 double
precision floating point type. This obviously adds some overhead to operations
on integers as overflow and non-integer divisions must be handled. However, this
is common practice in hand-written Javascript as well, and is generally handled
efficiently by Javascript engines.

Values of non-primitive data types in Haskell consist of a data constructor and
zero or more arguments. In Haste, these values are represented using arrays,
with the first element representing the data constructor and the following
values representing its arguments. For instance, the value \texttt{42 :: Int}
is represented as \texttt{[0, 42]}, the leading \texttt{0} representing the
zeroth constructor of the \texttt{Int} type and the \texttt{42}, obviously,
representing the ``machine'' integer. It may seem strange that a limited
precision integer is represented using one level of indirection rather than as
a simple number, but recall that the \texttt{Int} type is defined by GHC as
\texttt{data Int = I\# Int\#}.

@@

@@

While Conductance gets rid of the callback-based programming model endemic to
regular Javascript, it still suffers from many of its usual drawbacks. In
particular, the the weak typing of Javascript poses a problem in that the
programmer is in no way reprimanded by her tools for using server APIs
incorrectly or trying to transmit values which can not be sensibly serialized
and de-serialized, such as DOM nodes. Wrongly typed programs will thus crash, or
even worse, gleefully keep running with erroneous state due to implicit type
conversions, rather than give the programmer some advance warning that something
is amiss.

@@

Any and all communication between client and server is both strongly typed
and made explicit by the use of the \texttt{onServer} function, with the
programmer having complete control over the serialization and de-serialization
of data using the appropriate type classes. Aside from the obvious advantages
of type safety, making the crossing of the network boundary explicit helps the
programmer make an informed decision as to when and where server communication
is appropriate.

@@

\paragraph{Limitations} Unlike any of the related works, our approach takes a
firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client-side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Should this model prove
insufficiently expressive, extending our existing model with an
\texttt{onClient} counterpart to \texttt{onServer} would be relatively
straightforward.

@@

>>>>

\usepackage{amsmath}
\input{macros}

@@

\authorinfo{Koen Claessen and Anton Ekblad}
           {Chalmers University of Technology}
           {\{koen,antonek\}@chalmers.se}

@@

The runtime data representation of Haste programs is kept as close to regular
Javascript programs as possible. The numeric types are represented using the
Javascript \lstinline!Number! type, which is defined as the IEEE754 double
precision floating point type. This obviously adds some overhead to operations
on integers as overflow and non-integer divisions must be handled. However, this
is common practice in hand-written Javascript as well, and is generally handled
efficiently by Javascript engines.

Values of non-primitive data types in Haskell consist of a data constructor and
zero or more arguments. In Haste, these values are represented using arrays,
with the first element representing the data constructor and the following
values representing its arguments. For instance, the value \lstinline!42 :: Int!
is represented as \lstinline![0, 42]!, the leading \lstinline!0! representing
the zeroth constructor of the \lstinline!Int! type and the \lstinline!42!,
obviously, representing the ``machine'' integer. It may seem strange that a
limited precision integer is represented using one level of indirection rather
than as a simple number, but recall that the \lstinline!Int! type is defined by
GHC as \lstinline!data Int = I# Int#!.

@@

\paragraph{Interfacing with Javascript} While Haste supports the Foreign
Function Interface inherited from GHC, with its usual features and limitations
\ \cite{ffi}, it is often impractical to work within the confines of an
interface designed for communication on a very low level. For this reason
Haste sports its own method for interacting with Javascript as well, which
allows the programmer to pass any value back and forth between Haskell and
Javascript, as long as she can come up with a way to translate this value
between its Haskell and Javascript representations. Of course, not performing
any translation at all is also a valid ``translation'', which allows Haskell
code to store any Javascript value for later retrieval without inspecting it
and vice versa. The following example implements mutable variables using this
custom Javascript interface.

\begin{code}
import Haste.Foreign

-- A MutableVar is completely opaque to Haskell code
-- and is only ever manipulated in Javascript. Thus,
-- we use the Unpacked type to represent it,
-- indicating a completely opaque value.
newtype MutableVar a = MV Unpacked

instance Marshal (MutableVar a) where
  pack          = MV
  unpack (MV x) = x

newMutable :: Marshal a => a -> IO (MutableVar a)
newMutable = ffi "(function(x) {return {val: x};})"

setMutable :: Marshal a => MutableVar a -> a -> IO ()
setMutable = ffi "(function(m, x) {m.val = x;})"

getMutable :: Marshal a => MutableVar a -> IO a
getMutable = ffi "(function(m) {return m.val;})"
\end{code}

The core of this interface consists of the \lstinline!ffi! function, which
allows the programmer to create a Haskell function from arbitrary Javascript
code. This function exploits Javascript's ability to parse and execute
arbitrary strings at run-time using the \lstinline!eval! function, coupled with
the fact that functions in Haste and in Javascript share the same
representation, to dynamically create a function object at runtime.
The \lstinline!ffi! function is typed using a well-known type class trick for
creating statically typed variadic functions\ \cite{printf}, and works very
much like the \lstinline!printf! function of Haskell's standard library.
When applied to one or more arguments instantiating the \lstinline!Marshal!
type class, the \lstinline!pack! function is applied to each argument,
marshalling them into their respective Javascript representations, before they
are passed to the dynamically created function. When that function returns,
the inverse \lstinline!unpack! function is applied to its return value before
it is passed back into Haskell land.

As the marshalling functions chosen for each argument and the foreign
function's return value depends on its type, the programmer must explicitly
specify the type of each function imported using \lstinline!ffi!; in this,
Haste's custom method is no different from the conventional FFI.

There are several benefits to this method, the most obvious being that new
marshallable types can be added by simply instantiating a type class. Thanks
to the lazy evaluation employed by Haste, each foreign function object is only
created once and then cached; any further calls to the same (Haskell) function
will reuse the cached function object. Implementation-wise, this method is also
very non-intrusive, requiring only the use of the normal FFI to import
Javascript's \lstinline!eval! function; no modification of the compiler is
needed.

@@

While Conductance gets rid of the callback-based programming model endemic to
regular Javascript, it still suffers from many of its usual drawbacks. In
particular, the weak typing of Javascript poses a problem in that the
programmer is in no way reprimanded by her tools for using server APIs
incorrectly or trying to transmit values which can not be sensibly serialized
and de-serialized, such as DOM nodes. Wrongly typed programs will thus crash, or
even worse, gleefully keep running with erroneous state due to implicit type
conversions, rather than give the programmer some advance warning that something
is amiss.

@@

Any and all communication between client and server is both strongly typed
and made explicit by the use of the \lstinline!onServer! function, with the
programmer having complete control over the serialization and de-serialization
of data using the appropriate type classes. Aside from the obvious advantages
of type safety, making the crossing of the network boundary explicit helps the
programmer make an informed decision as to when and where server communication
is appropriate.

@@

\paragraph{Limitations} Unlike any of the related works, our approach takes a
firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client-side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Should this model prove
insufficiently expressive, extending our existing model with an
\lstinline!onClient! counterpart to \lstinline!onServer! would be relatively
straightforward.

While our solution completely safe from the invocation point of the
\lstinline!runApp! function onward, capturing IO-dependent variables introduced
before this invocation may lead to the client and server execution paths
diverging. Consider the following example:

\begin{code}
main = do
  captured <- randomRIO (False, True)

  runApp config $ do
    diverged <- export $ \transmitted -> do
      if captured == transmitted
        then return ``All OK''
        else return ``Diverged''

    runClient $ do
      divergence <- onServer (diverged <.> captured)
      alert divergence
\end{code}

The dialog box displayed by this program will read ``All OK'' roughly half
of the times it is run, while in half of the runs it will read ``Diverged''.
The reason for this is that the \lstinline!randomRIO! computation will be
executed twice - once on the client, and once on the server. Whenever the
computation returns differening results, the client and the server will have
conflicting opinions on the actual value of \lstinline!captured!, causing any
code which - quite reasonably - assumes that variables are consistent between
client and server to fail.

This is easily remedied however, by a slight compiler modification. Instead of
using \lstinline!main :: IO ()! as the program entry point, introduce a function
\lstinline!appMain :: App Done! for this purpose, and let the runtime system
start the application by invoking \lstinline!runApp config appMain!, where
\lstinline!config! contains the URL and port configuration for the server, given
by the programmer using a compiler flag. As this modification only affects the
program's entry point, which is already given ``special'' treatment in Haskell
compilers, we believe this to be a practical, non-intrusive way to remove this
limitation, and in fact made this modification to the Haste compiler for our
implementation.

The astute reader may notice that the same applies to code using
\lstinline!unsafePerformIO! to declare IO-dependent top-level values, and that
this problem is unaffected by changing the type of the program's entry point.
While this is indeed true, we feel that since it is quite possible to break
any Haskell program using \lstinline!unsafePerformIO!, even subverting the type
system, this is not a problem with our solution.

@@

\bibitem[Taylor (2013)]{printf}
C. Taylor: Polyvariadic Functions and Printf, http://chris-taylor.github.io/blog/2013/03/01/how-haskell-printf-works/, 2013
